---
title: "Syllabus EC438/538 Fall 2022"
author: "James Woods"
date: ''
output:
  pdf_document: default
  word_document: default
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Course Description

The official description of the course is:

> Economics and structure of energy markets, with a focus on electricity. Examines current policy issues arising from energy production and use.

This is a companion course to EC 437/537 which will address regulatory and competitive policies in electricity, public transportation, water, natural, and telecommunications. The order of the courses has changed to reflect the evolving nature of the field.



## Prerequisites

The undergraduate section has microeconomic theory, EC 311, or the calculus based version, EC 415, listed as prerequisites. The graduate section requires only graduate standing.


## Technology Requirements

```{r child = 'syllabus_common_files/tech_requirements.Rmd'}
```


## Health and Safety

While we are 'over' COVID-19, it does not mean that everything will return to the way it was.  I will continue to organize this course as an ``Attend Anywhere`` course with the option to join via zoom.  

If you are sick, not just COVID, please attend via zoom, or at least wear a mask.


# Contact Information

```{r child = 'syllabus_common_files/contact_info.Rmd'}
```

You can schedule one-on-one meetings at <https://woodsj.youcanbook.me/>. These will be online meetings through zoom.

# Basis for Grade




# The In-class Experience

My goal is to make

+ PERSON ON ZOOM IS ALWAYS SCRIBE



# Textbook and Other Resources

The main text for the course is Bhattacharyya, Subhes C. Energy economics: concepts, issues, markets and governance. Springer Nature, 2019..  It was chosen undergraduates in mind and illustrates many of the concepts they learned in EC 311/415 and EC 201 and focuses more on graphical treatments of the ideas.  

The book is intended to be a fairly comprehensive, mostly non-technical, overview of a sprawling and ragged field.

We will focus on a few core topics with the class voting on the remaining topics depending on interest and term paper topics.

Core Topics:

+ Introduction to Energy Economics (Ch 1): Brief introduction to the topic.
+ Energy Data and Energy Balance (Ch 2): Introduction to some units and energy balances.  Please note that the units are often oddly expressed in a mix of SI, US customary units and equivalents. Energy balance diagrams also have some quirks relating to how much energy is used to produce energy.
+ Energy Demand Analysis (Ch 3.1-3.6): This is the link to EC 311/415/201.  Please review your notes from these classes.  The final part of the chapter may be interesting to those that have taken Econometrics, EC469. 
+ Sectoral Energy Demand (Ch 4): We will use this chapter mostly as a gateway to learning about data sources.
+ Economic Analysis of Energy Investments (Ch 6): This is a brief version of part of EC 427 or EC 314. 

After we complete the core topics we will vote on the next topic to address based on student interest.  These topics will include outside readings that bring the topics more up to date and to emphasize that this is an upper-division economics course as opposed to a lower division or non-economics course.


Theory Options:  We should choose one of these

+ Energy Supply Development (Ch 7 and 8): This is about the decision to dig, drill, install solar panels and wind turbine using time value of money.

+ Energy Pricing and Taxation (Ch 9): Single period analysis 

Broad Topic Options:  These are multichapter options.  We will likely only address one of these with the majority of other topic determined by term paper choices.

+ Externalities and Climate Change (Ch 10, 11, 18, 19)

+ Markets (Ch 21, 22, 24, 25)

We can and will address other topics through our term papers.  Energy economics includes the effects of energy use too and many of you may be looking at the effects of electrification or fuel switching on education, health, property values.

## Writting Resources

I've put these resources on reserve in the library.  There are multiple, and more recent, editions of each of these books.  Don't get caught up on the specific edition.  All provide great guidance.  

McCloskey, D. (2000). Economical writing (Second ed.). Prospect Heights, Illinois: Waveland Press. is an excellent read on how to write. There are 31 short, snarky chapters in 91 pages. Read a chapter a day and try to apply what you read and you will be less embarrassed at work.

Chapter 19 in Wooldridge's, "Introductory Econometrics: A Modern Approach". It is a great outline of the process of empirical research.  Graduate students and those attempting an empirical paper should read this before choosing a topic for their term paper.

In addition to these resources I highly recommend using a tool, e.g., grammerly (https://www.grammarly.com), and stopping by the writing center (https://www.pdx.edu/writing-center/) before turning in drafts.


# Policy Statements from the University

```{r child = 'syllabus_common_files/psu_policy_statements.Rmd'}
```

Link to this syllabus https://github.com/woodsjam/Course-Energy-Economics.  


