---
title: "HW 1"
author: "Trailblazers"
date: "October 9, 2016"
output: pdf_document
---
``````{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

This is something like a notebook.  You can work with data and text at the same time.  Things chunks are for code
```{r}

print("I am in a chunk.")

```

By default they echo the code and then produce the result.

You can turn off the echo and just see the result.
```{r echo=FALSE}
print("No Echo")

```


```{r}
View(RealPrices)
summary(RealPrices)
```

## now plot Year and Real Price (test)

```{r}
plot(RealPrices$Year)
```
## now plot Year and short ton production
```{r}
plot(RealPrices$Year, CoalMarket$ProductionKShortTon)
```
## more plots
```{r}
plot(RealPrices$Year, RealPrices$RealPriceMBTU)
plot(RealPrices$Year, RealPrices$RealPriceMBTU)
plot(RealPrices$RealPriceMBTU, CoalMarket$ProductionKShortTon)
```
## Demand Curve
```{r}
plot(CoalMarket$ProductionKShortTon, RealPrices$RealPriceMBTU)
```
## now Multiple (linear) Regression
```{r}
FirstRegression<- lm(CoalMarket$ProductionKShortTon ~ RealPrices$RealPriceMBTU)
summary(FirstRegression)
plot(FirstRegression)
```
## now elasticities
FirstRegression<- lm(CoalMarket$ProductionKShortTon ~ RealPrices$RealPriceMBTU)
summary(FirstRegression)
plot(FirstRegression)


```{r}
ElastForm<- lm(log(RealPrices$Year) ~ log(RealPrices$RealPriceMBTU), data= RealPrices)
summary(ElastForm)
plot(ElastForm)
```
## Percent Form
```{r}
PercentForm <- lm(Year ~ log(RealPriceMBTU), data=RealPrices)
summary(PercentForm)
plot(PercentForm)
```
## Regression Real Prices and Production
```{r}
SecondRegression<- lm(CoalMarket$ProductionKShortTon ~ RealPrices$RealPriceMBTU)
summary(SecondRegression)
plot(SecondRegression)
