---
title: "Intro To Regression"
author: "James Woods"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment and Access 

1. Lets start with a few R conventions.  Lets start with scalars.  '<-' is the assignment operator in R.  '=' also works but '==' is used for comparisons.

```{r}
1+1
A <-  2+3
A + 2
B <- A +4
A+B
C = A+B
C
A == B
```

2. R has many ways of representing values
```{r}
A <- "Shale"
A
C <- c("Tom", "Dick", "Harry")
C

1:10
D <- 1:10
D
as.character(D)
as.factor(D)

summary(D)
summary(as.character(D))
summary(as.factor(D))

```

3. There are also functions. Most, but not all, work on many things at once.

```{r}
sqrt(D)
#sqrt(A)

sqrt(D[6])
sqrt(D[1:3])

```

4. If you need help, use the help pane or ask for help. Everything has a help file and many functions come with examples of how to use.

```{r}
help(sqrt)
example(sqrt)
```

4.  You can work with vectors and matrix but you will most frequently deal with dataframes, which is a matrix with extra attributes.  Dataframes are objects that have variables inside them.  You can access those variables with specific functions or with a '$'.

```{r}

X <- runif(20, 2, 40)
MyData <- as.data.frame(X)
MyData$Y <-  100 - 2*MyData$X +rnorm(20)

names(MyData)

names(MyData) <- c("Price","Quantity")
names(MyData)
summary(MyData)
```

5.  You can get at columns and rows in other ways.

```{r}

MyData[1:2]
MyData[1]
MyData[1:3,]
MyData[1:3,"Price"]
MyData["Price"]
MyData[c(1,5,8),]
```

# Pictures
Now lets make pictures.
```{r}
plot(MyData$Price)
```

Historgram
```{r}
hist(MyData$Price)
```

Show the relationship
```{r}
plot(Quantity~Price, data= MyData)

```

# Regression with Pictures

5. A regression line can be thought of as just putting a line through a cloud of data in a well defined way.  Remember the data wer created.
```{r}
FirstRegression<- lm(Quantity ~ Price, data=MyData)

summary(FirstRegression)
```

Note that the estimates are the parameter, or very close to the parameters we used to create the data.  Note also that the residual standard errors are the same as what we used to create the data.  Nice when you know what is true.

Regression can also show you things that are not true. You have to assume a functional form.  Make new data.


```{r}

MyData2<- data.frame(Price = runif(20, 2, 40))
MyData2$Quantity <-  100 - 2*MyData2$Price + .05*MyData2$Price^2 +rnorm(20)

summary(MyData2)
```

Not a line
```{r}
plot(Quantity~Price, data= MyData2)

```

Give me a line anyway.

```{r}
NotALine<- lm(Quantity ~ Price, data=MyData2)

summary(NotALine)
```

How to tell you were wrong thinking it was a line. Make a pictures
```{r}
plot(NotALine)

```

Compare with when it was a line and you estimated a line

```{r}
plot(FirstRegression)

```

You can make these by hand too.  Here are two accessor functions.

```{r}
plot(resid(NotALine)~fitted(NotALine))

```


# Dummy Variables

# Looking at Residuals for problems



# Real Data

2. I will  walk you through a few steps on reading in the data. The biggest hurdle to doing stats on the data is reading it in. There is actually an R library for working with EIA data directly, EIAdata, but we will use the more general tools to read files.



3. Download into R data on coal prices and quantities.  Again, the assignment operator in R is the "<-" symbol.

```{r}

Coal <- read.csv("https://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T06.01")

```

There are many ways of loading data into R (http://www.r-tutor.com/r-introduction/data-frame/data-import).  Some work some of the time.  In most cases Comma Separated Values (CSV) is the safest format to work with.

4. Take a look at the summary of the data

```{r}
summary(Coal)

```

You will notice that for some variables they give counts, e.g., MSN, and for others you get numerical summaries, e.g., Column_Order.  The difference has to do with the data types (http://www.statmethods.net/input/datatypes.html).

5.  Since we are trying to make a simple supply model, i.e., trying to explain coal production, lets select just the production part of the data set and also get only the annual production values.  This is a little primer on changing data types and taking part of data.

+ Load the dplyr package.  This is the normal way to load libraries of functions that you need.
```{r}
library(dplyr)

```

+ Select only the Coal Production Figures and save it as CoalProduction.  There is a cheat sheet for dplyr built into R.  Look under the help menu.
```{r}
CoalProduction <- Coal %>% filter(MSN == "CLPRPUS") 

```

+ Note that you have monthly data but that the annual data is shown as month 13.  Grab all of the observations with month equal to 13.

```{r}
library(stringr)

CoalProduction <- CoalProduction %>% filter(str_sub(as.character(YYYYMM),5 ) == "13")

```
+ At this point you will have noticed that we don't need all the columns so lets keep just the ones we need and then give the two columns we saved new names.

```{r}
CoalProduction <- CoalProduction %>% select(YYYYMM, Value)

names(CoalProduction) <- c("RawYear", "ProductionKShortTon")
summary(CoalProduction)
```
+ Notice that the ProductionKShortTon variable shows a count rather than a numerical summary.  This means that R thinks it is a factor rather than a number.  Lets fix that.  It requires to first convert the factor, which is an integer, to the real value as a character and then convert that to numeric.

```{r}
CoalProduction$ProductionKShortTon <- as.numeric(as.character(CoalProduction$ProductionKShortTon))

```
Play around with this doing one function at a time to see what each does and what each does alone.

+ Next lets create a column for the year and make it a numeric value.
```{r}

CoalProduction <- CoalProduction  %>% mutate(Year = as.numeric(str_sub(as.character(RawYear),0,4 )))


summary(CoalProduction)
```

5. Now grab some price data.  We are going to use Quandl (https://www.quandl.com/), which has a bunch of data on energy and a lot of other things (https://www.quandl.com/collections/markets/coal). Make sure you have the Quandl library installed. If you don't install.packages("Quandl") should do it.  

```{r}
library(Quandl)
Prices <- Quandl("EPI/152")

summary(Prices)

```
+ As before we will convert the year to a numeric value
```{r}
Prices$Year <- as.numeric(str_sub(Prices$Year,0,4))


```
+ And simplify the names
```{r}
names(Prices) <- c("Year", "PriceMBTU")

```

Please note that we don't have the price per short ton of coal.  What we have is the price per million BTUs, which is a measure of energy content.  The BTUs per short ton of coal (2000 lbs) is about 20 MBTUs but varies from place-to-place and year-to-year.

+ Merge the two data frames

```{r}
summary(CoalProduction)
summary(Prices)
CoalMarket<-inner_join(Prices, CoalProduction, by ="Year")

summary(CoalMarket)

```
