---
title: "Reading Regression Output Without Taking Econometrics"
author: "James Woods"
date: ''
output:
  beamer_presentation: default
  

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



## Just Basics

Doing Econometrics well is hard.

+ EC 460 will teach you the basics of single-equation regression, e.g., demand estimation, and how to fix some things.
+ Make you overconfident.
+ Functional work requires:
    + Systems equation estimation, e.g., supply and demand at the same time.
    + Discrete choice, e.g., Yes/No or make and model of a car. 
    + Knowledge of how to get causality, experimental, e.g., RCT, and quasi-experimental methods, e.g., regression discontinuity.
    
## BASICS Single Equation

Explain something, often called the left-hand side or endogenous variable, with explanatory, right-hand side or exogenous variables.

$$Weight = \alpha + \beta Height + \epsilon$$

+ Weight is left-hand side
+ $\alpha$ intercept term, expected weight given you have no height. 
+ $\beta$ How much your weight increases for every inch of height.
+ $\epsilon$ How far off we were.
+ All the greek letters are random variables.  We estimate the means and variances of those.

Generating Fake Data with Weight = 20 + 2.75 Height + n(0,40)

## Fake Data

```{r echo=FALSE, message=FALSE, warning=FALSE}


HWFake<- data.frame(Height = rnorm(mean = (12*5)+10, sd = 10, n= 100))
                    
HWFake$Weight <- 20 + rnorm(mean = 20, sd = 40, n = 100) + 2.75*HWFake$Height

library(ggplot2)
library(dplyr)

HWFake %>%
  ggplot(aes(x = Height, y = Weight)) +
  geom_point() 
  
```


## Regression Output

```{r echo=FALSE, message=FALSE, warning=FALSE , results = "asis"}
library(stargazer)

stargazer(lm(Weight ~ Height, HWFake), header = FALSE )

```

## Parts

Regression as a whole:

+ $R^2$ fraction of variation in LHS explained by variation in RHS.
    + Because if you add more variables $R^2$ goes up, Adjusted $R^2$, penalizes for having more variables.
+ $F$ Difference between model that is mean of LHS alone, just the intercept, vs model with RHS.


The parameters:

+ 
    