---
title: "Energy Models"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Overview of Energy Models
  + Trend
  + Time Series
  + Guess based on experience
  + Survey Based
  + Scenario Planning
  + I/O Models
  + Energy Balance Models
  + Game Theory
  + Experimental/Behavioral

# Trend
  + Uses only patterns of the past to make forecasts about the future.
  + Cram a line though the data.
  
  
```{r Trend, echo=FALSE, message=FALSE,fig.width=6, fig.height=4}
library(dplyr)
Coal <- read.csv("https://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T06.01")
CoalProduction <- Coal %>% filter(MSN == "CLPRPUS") 
library(stringr)
CoalProduction <- CoalProduction %>% filter(str_sub(as.character(YYYYMM),5 ) == "13")
CoalProduction <- CoalProduction %>% select(YYYYMM, Value)
names(CoalProduction) <- c("RawYear", "ProductionKShortTon")
CoalProduction$ProductionKShortTon <- as.numeric(as.character(CoalProduction$ProductionKShortTon))
CoalProduction <- CoalProduction  %>% mutate(Year = as.numeric(str_sub(as.character(RawYear),0,4 )))
library(ggplot2) 
qplot(Year, ProductionKShortTon, data = CoalProduction) + geom_smooth(method= lm)

```

# Trend Comments
+ General 
    + Regression with time as RHS variable. $Y = time + \epsilon$
    + Other variables are often added, for example, weather variables.  You can see this in some integrated resource planning models.
    + EC469 shows you how to do this.
    + Not everything is a line.  You sometimes have to transform the data, logs and such are common.

+ Problems
    + Only uses the past and does not show turns in trend.
        + I was 90 lbs at 15, 110 at 17, 130 at 18 and 220 at 26.
        + Today?
    + People don't react to prices and there is no change in technology. 
        + See the coal data at the end.  That is the natural gas fracking boom.
    + Growth of a subset can only exceed that of a superset for a while.  
        + Facebook can only grow faster than the US, or world, economy for so long.

# Trend Comments (Con't)

+ Advantages
    + That said, it works for a lot of things and work when you have little time.
    + Plenty of energy modeling looks like this.
        + "If you don't have the time, throw in a trend line."
    + For looking a few periods into the future -- it works.  For longer term forecasts -- don't do it.
    
# Time Series

+ Similar to trend only
    + May have $Y  = ARMA(p,q)$
    + or include a trend $Y = time + ARMA(p,q)$
    + the AR part is past values of LHS
    + the MA part is past values of error term.
+ EC 472 shows you how to do this.
+ Most energy data has a trend to it, which must be included in the model.
+ Best thought of, for now, as a refinement to the trend regressions with better treatment of residuals.
    + More accurate confidence intervals on existing data.
    + Slightly better with near-term forecasts.
+ Similar problems as pure regression on trend.


# There is way more to time series

+ Macro is heavy into this.
+ You may see macro heavy energy papers with Vector Autoregressions (VARs) which looks at many time series at once.
+ Nice example of quick use by Borenstein about the California Cap and Trade (https://ei.haas.berkeley.edu/research/papers/WP281.pdf)
    + They produced the key result that the price will not be at a bound, upper or lower, only 19% of the time.
    + Quick analysis.

# Guess based on experience

+ Don't laugh, this works.
+ Old hands make very good guesses based on experience.
+ GDP even has components that are based on analyst's best judgement.
+ This can be combined with incentivised polls or parimutuel betting to make it work better.
  + The Iowa Electronic Markets are a common example, though there are others. https://tippie.biz.uiowa.edu/iem/markets/ 

# Incentivized Polls

+ Guess Jamie's BMI.  Closest to the actual gets $10.
+ Skin in the game gives incentive for more thought.
+ Only works when you will eventually find out actual.
+ Oddly works better if you don't know how people are betting.
    + Can produce group-think bubbles, if you see how others are betting.
    + Easy to find the mode, but harder to tell if it is real.
+ Can get expensive with many winners:
    + everyone guesses the same correct answer.
    + More expensive when there is less certainty.
  

# Parimutuel Betting

+ Guess Jamie's BMI. Correct answers split the pot with the others that got it right.
+ Racetrack method.
+ Less likely to get bubbles, but
    + Harder to see if there is a mode in the guess.
+ Cheaper than incentivized poll if high probability of multiple winners.
+ Cost does not increase with a priori certainty.


# Survey Based

+ Book gives some examples.
+ Ask a bunch of experts about their best judgement and summarize
+ Often expanded as a Delphi Survey
+ For comically bad forecast see "Results of the Delphi IX Survey of Oil Forecasts" California Energy Commission, 1997.

# Delphi Surveys

+ Ask each person _in private_ for best guess.
+ Compile results.
+ Ask outliers why they said what they did.
+ Give everyone:
    + The distribution of guesses for each parameter
    + The reasons the outliers gave for the answer the gave.
    
+ Ask for another guess.
+ Report the new distribution or repeat if desired.


# Scenario Planning

+ Too much variation to give one description. 
+ General Steps:
    + Choose some parameters of interest, e.g., environmental activism, sea level rise, fracking shown to cause earthquakes, etc.
    + Package them into narratives that are possible.  Some improbable and some likely.
    + Decide how you would act now given that you know that that future will exist.
    + ?
    + Profit
+ Example by National Renewable Energy Research Laboratory (NREL) did one for Hawaii. http://www.nrel.gov/docs/fy12osti/52442.pdf
    + Contractor reports tend to go long.
    + See page 6 for summary of the scenarios.
    + Contractor reports tend to use lots of acronyms.
    + Plug in Hybrid Electric Vehicle (PHEV)
    

# Frequent use and abuse of scenarios analysis

+ Rigging scenarios so that only one is taken seriously. (abuse)
+ Choosing actions that will "work" in all scenarios. (abuse)
+ Actual plans if a scenario becomes more likely, "Plans are worthless, but planning is everything." - Ike
+ Creation of real options
    + Defer action until more information is available.  Don't bring an umbrella everywhere, but have on near by if it starts to rain.
    + Create systems you can use when the situation is clearer.  Think peaking power plants that only run a few days a year.

# I/O Models
+ Old school -- 1920s Leontief
+ Has an equilibrium concept
+ Assumes fixed ratios are used in production
    + No reaction to price changes
    + No reaction to input price changes
    + Constant returns to scale.
+ Often seen as part of a computable general equilibrium model to shorten run-times. REMI (http://www.remi.com/) and IMPLAN (http://implan.com/) use it in regional economic models.

+ Will not ask you to do one unless you want to.
+ Book has an overly long explanation.

# Walk Through I/O model
$x_1 = \alpha_{1,1} x_1 + \alpha_{1,2} x_2 + d_1$

+ $x_1$ is how much of good one that gets made.
+ $d_1$ is how much final consumers want of good 1.
+ $\alpha_{1,1}$ the amount of good 1 needed to produce good 1.
+ $\alpha_{1,2}$ the amount of good 2 needed to produce good 1.
+ Each good has an equation
+ $\alpha$s can be zero but there are restrictions on how many and where. The matrix needs to be invertable.

Matrix Form
$x=Ax +d$ is solved as $x=(I-A)^{-1}d$

d is what you need to fight the war.

# Warnings about the next models in book

+ They are an unholy mix of estimates, in engineering sense, and estimates, in the statistical sense.
+ They can be huge.  The national energy modeling system has many modules and sub-modules.
    + http://www.eia.gov/forecasts/aeo/nems/overview/electricity.html
+ The approaches are frequently combined.

# "Estimate" to an Engineer and Statistician

+ To an engineer, an estimate means taking some known or assumed values for parameters then performing some calculations to find a parameter of interest.
    + Example, building energy simulations like EnergyPlus.  Put in windows, screens, walls, assume some values human behavior, and weather and out pops annual energy use.
    + Can work with _very_ complex models, but is usually shaky with uncertainty.
+ To a statistician, an estimate means taking data, repeated observations of a physical data generating process and calculating a summary statistic.
    + Example, estimate implied thermostat settings from gas use and R values of walls from a survey.
    + Treats uncertainty well but often lack the complexity of the engineering style models.
    
# Combining Engineering and Statistician Style Estimates

+ A typical Example is estimates of  Unit Energy Consumption (UECs)
    + The EIA Residential Energy Consumption Survey (RECS) http://www.eia.gov/consumption/residential/reports/2009/methodology-end-use.cfm
    + California Residential Appliance Saturation Study (RASS) http://www.energy.ca.gov/appliances/rass/
    
+ Check the details for RASS and see how they combine engineering estimates with regression analysis.     
    + http://www.energy.ca.gov/2010publications/CEC-200-2010-004/CEC-200-2010-004-V1.PDF.
    + Often called a statistically adjusted engineering model.

# Game Theory

+ Not a full, all economy, model but a tool used to deal with decisions where:
    + There is not a monopoly or monopsony.
    + Not perfect competition.
+ Also used for:
    + Dynamic interactions of firms, think how gasoline prices go up fast but down slow.
    + Auction and bidding, technically mechanism design which is game theory backwards, to get people to tell the truth or do the right thing.
    + Basis for a lot of modern "regulation" which focuses on encouraging competition to reach goals.
+ Probably did some in EC 201 or EC 311, but we have a full class EC 321 and math has several.
    

# Experimental/Behavioral

+ The behavioral part is admitting that people and firms do not act rationally.
+ Old school experiments, A/B testing and more, run on people
+ Field experiments are common in economics now
    + Esther Duflo received the Bates Clark Award 2010
    + Bluffstone in Econ is running two now in Ethiopia and Nepal.
    
+ Typical Issues
    + Internal validity
        + Can you really connect cause to effect?
        + Did you avoid bias and control for everything?
    + External validity
        + Does it work in real life?
        + Does it work on other people?
    + Ecological validity
        + Did the experiment look like the real world?
        
# Overarching

+ Every model is wrong
    + By definition they are simplifications of reality and they leave out details.
    + The real question is, "Did they leave out the right details?"
+ Don't be deluded into, "Just look at the data".
    + You come to data with a model, you just don't know it.
    + The data never speak for themselves. You use data to test/support/refute hypotheses/theories.
+ Using more than one model or technique and getting the same general result helps. It shows that your conclusion about reality is resilient to the implicit modeling assumptions.
+ Keep in mind that modeling is an iterative process often driven by the researcher.
    + You try a model and then refine it based on the results.
    + It is an iterative process.
    + Your choices are molded by your objectives. "Get published" is common.
    + Great working paper on this by Gelman that expands on this idea "The garden of forking paths: Why multiple comparisons can be a problem,
even when there is no “fishing expedition” or “p-hacking” and the research hypothesis was posited ahead of time" (http://www.stat.columbia.edu/~gelman/research/unpublished/p_hacking.pdf)